When does vapor pressure deficit drive or reduce evapotranspiration?

### 2017-10-23 doc/paper/vpd_et_paper.tex
some writing changse

Changes to vapor pressure deficit (VPD) alter the atmospheric demand for water from the land surface. Traditionally, atmospheric scientists and hydrometeorologists generally think that an increase in atmospheric demand induces an increases in evapotranspiration (ET) (citations?). This possible misconception developed in part due to the proliferation of studies examing potential ET rather than estimates of ET itself (citations?).  In contrast, plant physiologists know that stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase (decrease) in VPD may not correspond to an increase (decrease) in ET because stomatal closure (opening) can cancel the effects of shifts to atmospheric demand.

@@

@@

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so GPP in \ref{uwue} is really $A\cdot \text{LAI}$. We also recognize that our use of $uWUE$ introudces considerable uncertainty, so we added an uncertainty parameter $\sigma = f(t, \text{site})$ modifying $uWUE$. This will be elaborated on below.

@@

\begin{linenomath*}
  \begin{equation}
  ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p D_{s}}{R_{air}} -  \frac{\gamma c_s \sqrt{D} }{ R* \; 1.6\; \sigma\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D}})} \right) }{ \Delta + \gamma}
  \end{equation}
\end{linenomath*}

We can then take the derivative with respect to $D$ to analytically determine ecosystem response to atmospheric demand perturbations:

@@

%% see registers!!!!!! - help for emacs productivity
This $\sigma$ will be some part uncertainty associated with assumptions about constant $g1$ and $uWUE$, including differences between the data used here and in the fits in \cite{Zhou_2016} and \cite{Lin_2015}, but also some general PM model and FLUXNET observationsal uncertainty. By calculating a unique $\sigma$ for each observation we will propagate any model and observational uncertainty forward into our expression for $\frac{\partial \; ET}{\partial D}$.

@@

By construction, the variability in the $\sigma$ term (Equation \ref{lai}) contains all model and observational uncertainties. We expect LAI to be cloase to LAI$_{ref}$, so $\sigma$ should be $O(1)$. We can have some confidence in our framework, including the assumption of constant uWUE, if calculated $\sigma$s are generally $O(1)$. Figure \ref{lai_fig} presents the histogram of calculated $\sigma$s with outliers (lowest and highest 5\% percent) and nonphysical values ($\sigma$ $<$ 0.) removed. All remaining $\sigma$ values are $O(1)$ which provides confidence in model framework.

@@

An additional concern is that the $\sigma$ term may in fact be some function of $D$, in which case the dependence would need to be accounted for when taking the derivative. Figure \ref{lai_vpd_fig}\explain{Figs 2 and 3 can probably be combined - the histogram of $\sigma$ is shown in Fig 3}. plots the joint distribution of $\sigma$ and VPD, and shows that $\sigma$ is very weakly a function of VPD. Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $D$.

@@

Before calculating the sensitivity of ET to VPDn, it is useful to consider the functional form of Equation \ref{d_et}. There are three terms: a scaling term for the full expression we will call Term 1 ($\frac{g_a \; P}{T(\Delta + \gamma)}$), a relatively constant offset we will call Term 2 ($\frac{c_p}{R_{air}}$), and a variable term we will call Term 3 ($\frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right)$). All variables are positive, so the relative magnitude between Term 2 and Term 3 will determine the sign of the derivative, while Term 1 will scale the expression larger or smaller.

@@

Figure \ref{scale_vary}A shows Term 1 normalized by mean $g_a$ (calculated for each plant functional type), and confirms that much of the relative variability of Term 1 is contained in the $g_a$ term's relative variability. Generally, $T$ has less of a role. Additionally, the impact of $T$ on the relative variability increases with increasing $g_a$. % plot every PFT and show they collapse onto the same vurve?

@@

Term 2 minus Term 3 determines the sign of the sensitivty of ET to VPD and is thus crucial. $c_s$ variability is relatively less than $\sigma$ and $D$ variability, then variability within PFT will be solely determined by $\sigma$ and $D$ at the different fluxnet sites. Figure \ref{term3} shows how (Term 2 - Term 3) varies with $D$ and $\sigma$, as a function of PFT. In Figure \ref{term3}a lower uWUE and $\sigma$ shift the distribution of (Term 2 - Term 3) towards negative values. Additionally, the smaller $g1$, the greater the relative $D$ dependence of of (Term 2 - Term 3). This is observed most strongly for the ENF PFT, which has the smallest $g1$ (74.31).

@@

The intuition developed using this framework can be used to understand how the land surface will respond and contribute to changes in the environment.

@@

%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
%
% \begin{thebibliography}{}
% \bibitem[{\textit{Kobayashi et~al.}}(2003)]{R2013} Kobayashi, T.,
% Tran, A.~H., Nishijo, H., Ono, T., and Matsumoto, G.  (2003).
% Contribution of hippocampal place cell activity to learning and
% formation of goal-directed navigation in rats. \textit{Neuroscience}
% 117, 1025--1035.
%
% \bibitem{}
% Text
% \end{thebibliography}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Or, to use BibTeX:
%
% Follow these steps
%
% 1. Type in \bibliography{<name of your .bib file>} 
%    Run LaTeX on your LaTeX file.
%
% 2. Run BiBTeX on your LaTeX file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Run LaTeX on your new file which will produce the citations.
%
% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.

\input{vpd_et_paper.bbl}
%\bibliography{references.bib}

>>>>

Changes to vapor pressure deficit (VPD) alter the atmospheric demand for water from the land surface. Traditionally, atmospheric scientists and hydrometeorologists generally think that an increase in atmospheric demand induces an increases in evapotranspiration (ET) (citations?). This possible misconception developed in part due to the proliferation of studies examing potential ET (PET) rather than estimates of ET itself (citations?).  In contrast, plant physiologists know that stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase (decrease) in VPD may not correspond to an increase (decrease) in ET because stomatal closure (opening) can cancel the effects of shifts to atmospheric demand.

@@

We hypothesize that for most plant types a common resonse to increase in VPD will actually be a decrease in ET. The excpetion would be plants such as crops that are evolved (or bred) to prioritize gross primary production (GPP) over water conservation. However, for all otehr plants types, our hypothesis calls into question the validity of PET-based drought metrics deceloped by hydrometeorologists and used extensively in operations \citep[e.g. PDSI, P-PET,][]{2002, Alley_1984} \explain{more citations needed, including recent PET climate studies like Jack Scheff}. These metrics ignore the role of plants as gatekeepers for surface water loss to the atmosphere and have limited physical meaning for drought of vegetated land types. Additionally, plants evolved in arid climates should prioritize water conservation and we would expect a very negative ET response to increase in VPD. Therefore, vegetated locations most likely to experience droughts should show the strongest deviation between reality and a PET-based approximation.

@@

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so GPP in \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancelation of LAI in additoin to uWUE in Equation \ref{medlyn}. We also recognize that our use of $uWUE$ introudces considerable uncertainty, so we added an uncertainty parameter $\sigma = f(t, \text{site})$ modifying $uWUE$. The meaning of this parameter will be elaborated on after the derivation of $\frac{\partial \; ET}{\partial \; D}$.

@@

\begin{linenomath*}
  \begin{equation}
    ET = \frac{\Delta R + \frac{g_a\; P}{T} \left( \frac{ c_p D_{s}}{R_{air}} -  \frac{\gamma c_s \sqrt{D} }{ R* \; 1.6\; \sigma\; \text{ uWUE } (1 + \frac{g_1}{\sqrt{D}})} \right) }{ \Delta + \gamma}
    \label{et}
  \end{equation}
\end{linenomath*}

Note that if we set the uncertaitny parameter $\sigma$ to 1, and approximage $c_s$ at a global mean CO$_2$ concentration, then Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015, Lin_2015}. This then begs the question, why use PET at all for drought metrics in vegetated areas? A much more physically realistic estimate of ET can be had with the same information reqiured to calculate PET.

We can now take the derivative with respect to $D$ to analytically determine ecosystem response to atmospheric demand perturbations:

@@

This $\sigma$ will be some part uncertainty associated with assumptions about constant $g1$ and $uWUE$, including differences between the data used here and in the fits in \citet{Zhou_2015} and \citet{Lin_2015}, but also some part general PM model and FLUXNET observationsal uncertainty. By calculating a unique $\sigma$ for each observation we will propagate any model and observational uncertainty forward into our expression for $\frac{\partial \; ET}{\partial D}$.

@@

By construction, the variability in the $\sigma$ term (Equation \ref{lai}) contains all model and observational uncertainties. For an observation that perfectly matches our model and assumptions $\sigma$ will be one. Therefore, if for our assumptions and framework to be reasonable $\sigma$ should be $O(1)$. Figure \ref{lai_fig} presents the histogram of calculated $\sigma$s with outliers (lowest and highest 5\% percent) and nonphysical values ($\sigma$ $<$ 0.) removed. All remaining $\sigma$ values are $O(1)$ which provides confidence in model framework.

@@

An additional concern is that the $\sigma$ term may in fact be some function of $D$, in which case the dependence would need to be accounted for when taking the derivative. Figure \ref{lai_vpd_fig}\expplain{Figs 2 and 3 can probably be combined - the histogram of $\sigma$ is shown in Fig 3.} plots the joint distribution of $\sigma$ and VPD, and shows that $\sigma$ is very weakly a function of VPD. Given this weak dependence, we argue that Equation \ref{d_et} is a valid approximation for ET response to $D$.

@@

Before calculating the sensitivity of ET to VPD, it is useful to consider the functional form of Equation \ref{d_et}. There are three terms: a scaling term for the full expression we will call Term 1 ($\frac{g_a \; P}{T(\Delta + \gamma)}$), a relatively constant offset we will call Term 2 ($\frac{c_p}{R_{air}}$), and a variable term we will call Term 3 ($\frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{D}}{2 (g_1 + \sqrt{D})^2}\right)$). All variables are positive, so the relative magnitude between Term 2 and Term 3 will determine the sign of the derivative, while Term 1 will scale the expression larger or smaller.

@@

Figure \ref{scale_vary}A shows Term 1 normalized by mean $g_a$ (calculated for each plant functional type), and confirms that much of the relative variability of Term 1 is contained in the $g_a$ term's relative variability. Generally, $T$ has less of a role. Additionally, the impact of $T$ on the relative variability increases with increasing $g_a$. \explain{plot every PFT and show they collapse onto the same curve? - might just be too messy though}.

@@

Term 2 minus Term 3 determines the sign of the sensitivty of ET to VPD and is thus crucial. $c_s$ variability is relatively less than $\sigma$ and $D$ variability, so variability within PFT will be solely determined by $\sigma$ and $D$ at the different fluxnet sites. Figure \ref{term3} shows how (Term 2 - Term 3) varies with $D$ and $\sigma$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE$, $\overline{\sigma}$, and $g1$ shift the leading-order constant term (\frac{1}{g_1}) towards smaller values, and (Term 2 - Term 3) towards positive values. $uWUE$ and $g1$ are both water-use effiency type constants. Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water. Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3, while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exerts a greater influence than difference $uWUE$, as the power of $g1$ increases. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect due to its increasing powers. ENF ($g1 = 74.31$) has by far the VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.

@@

The intuition developed using this framework can be used to understand how the land surface will respond and contribute to changes in the environment. Additionally, Equation \ref{et} gives provides an estimate of ET that requires no additional information beyond that required to calculate PET. Given that for all PFTs, with the exception of CRO, we found a high frequency and magnitude of negative $\frac{\partial \; ET}{\partial \; D}$, PET is a physically unrealisitic representation of ET for vegetated surfaces and PET-based drought metrics are not usefull. We advocate for drought metrics using Equation \ref{et} instead of PET.

@@

%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
%
% \begin{thebibliography}{}
% \bibitem[{\textit{Kobayashi et~al.}}(2003)]{R2013} Kobayashi, T.,
% Tran, A.~H., Nishijo, H., Ono, T., and Matsumoto, G.  (2003).
% Contribution of hippocampal place cell activity to learning and
% formation of goal-directed navigation in rats. \textit{Neuroscience}
% 117, 1025--1035.
%
% \bibitem{}
% Text
% \end{thebibliography}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Or, to use BibTeX:
%
% Follow these steps
%
% 1. Type in \bibliography{<name of your .bib file>} 
%    Run LaTeX on your LaTeX file.
%
% 2. Run BiBTeX on your LaTeX file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Run LaTeX on your new file which will produce the citations.
%
% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.
\bibliography{references.bib}
\input{vpd_et_paper.bbl}
